---
title: "Exploratory Biomarker Analyses, BRAIN-ICU Study"
author: 'Jennifer Thompson, MPH; Supervisor: Rameela Chandrasekhar, PhD'
date: "September 18, 2015"
output: html_document
---

```{r options, echo=FALSE, results='hide'}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, results = 'hide')
```

```{r setup}
if(Sys.info()['sysname'] == 'Darwin'){
  load('/Volumes/thomps23/ICUDelirium/BRAINICU/braindata.Rdata')
} else{
  load('/home/thomps23/ICUDelirium/BRAINICU/braindata.Rdata')
}

biomarker.list <- c('crp', 'ifngamma', 'il10', 'il12', 'il1b', 'il6', 'il8', 'mmp9', 'proteinc', 'tnfa', 'tnfr1')

library(dplyr)
library(tidyr)
library(Hmisc)
library(caTools)
library(pROC)

## -- Function to round and format numeric value to same number of places -- ##
round.format <- function(quant, dig = 0){
  format(round(quant, digits = dig), nsmall = dig)
}

## -- Function to create string of descriptive statistics for plots -- ##
get.results.string <- function(qn, q50, q25, q75, qmean, qsd){
  gsub(' +', ' ',
       paste0('N: ', qn, ' (', round((qn / nrow(brain.oneobs))*100),
         '%)\nMedian (IQR): ', round.format(q50, dig = 2),
           ' (', round.format(q25, dig = 2), ', ', round.format(q75, dig = 2),
         ')\nMean (SD): ', round.format(qmean, dig = 2), ' (', round.format(qsd, dig = 3), ')'))
}

# Multiple plot function: Winston Chang, R Cookbook
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

We present descriptive statistics and data visualizations for the `r length(biomarker.list)`
biomarkers collected in both the BRAIN-ICU and MIND-ICU cohorts. For speed and ease, only BRAIN-ICU
data is included here, assuming that patterns will be similar for MIND-ICU.

```{r descstats}
## -- Calculate mean, peak, minimum, change, and AUC of each biomarker for each patient -- ##
bio.summary.pt <- brain.daily[,c('id', 'study.day', paste0(biomarker.list, '.imp'))] %>%
  gather(key = bio.marker, value = bio.value, crp.imp:tnfr1.imp) %>%
  filter(!is.na(bio.value)) %>%
  arrange(id, study.day) %>%
  group_by(id, bio.marker) %>%
  summarise(pt.n = sum(!is.na(bio.value)),
            pt.mean = mean(bio.value, na.rm = TRUE),
            pt.min = min(bio.value, na.rm = TRUE),
            pt.max = max(bio.value, na.rm = TRUE),
            pt.first = head(bio.value, n = 1),
            pt.last = tail(bio.value, n = 1),
            pt.auc = trapz(x = study.day, y = bio.value))
bio.summary.pt$pt.delta <- with(bio.summary.pt, ifelse(pt.n == 1, NA, pt.last - pt.first))

## -- Summary statistics for each biomarker across all patients; create strings to put on plots -- ##
## Spaghetti plots: % (number) of patients with 1, 2, 3 biomarker measurements
## Min, max, mean, delta, AUC: N, median (IQR), mean (SD)
bio.summary <- bio.summary.pt %>%
  group_by(bio.marker) %>%
  summarise(n.1 = sum(pt.n == 1, na.rm = TRUE),
            n.2 = sum(pt.n == 2, na.rm = TRUE),
            n.3 = sum(pt.n == 3, na.rm = TRUE),
            mean.n = sum(!is.na(pt.mean)),
            mean.25 = quantile(pt.mean, probs = 0.25, na.rm = TRUE),
            mean.50 = quantile(pt.mean, probs = 0.50, na.rm = TRUE),
            mean.75 = quantile(pt.mean, probs = 0.75, na.rm = TRUE),
            mean.mean = mean(pt.mean, na.rm = TRUE),
            mean.sd = sd(pt.mean, na.rm = TRUE),
            min.n = sum(!is.na(pt.min)),
            min.25 = quantile(pt.min, probs = 0.25, na.rm = TRUE),
            min.50 = quantile(pt.min, probs = 0.50, na.rm = TRUE),
            min.75 = quantile(pt.min, probs = 0.75, na.rm = TRUE),
            min.mean = mean(pt.min, na.rm = TRUE),
            min.sd = sd(pt.min, na.rm = TRUE),
            max.n = sum(!is.na(pt.max)),
            max.25 = quantile(pt.max, probs = 0.25, na.rm = TRUE),
            max.50 = quantile(pt.max, probs = 0.50, na.rm = TRUE),
            max.75 = quantile(pt.max, probs = 0.75, na.rm = TRUE),
            max.mean = mean(pt.max, na.rm = TRUE),
            max.sd = sd(pt.max, na.rm = TRUE),
            delta.n = sum(!is.na(pt.delta)),
            delta.25 = quantile(pt.delta, probs = 0.25, na.rm = TRUE),
            delta.50 = quantile(pt.delta, probs = 0.50, na.rm = TRUE),
            delta.75 = quantile(pt.delta, probs = 0.75, na.rm = TRUE),
            delta.mean = mean(pt.delta, na.rm = TRUE),
            delta.sd = sd(pt.delta, na.rm = TRUE),
            auc.n = sum(!is.na(pt.auc)),
            auc.25 = quantile(pt.auc, probs = 0.25, na.rm = TRUE),
            auc.50 = quantile(pt.auc, probs = 0.50, na.rm = TRUE),
            auc.75 = quantile(pt.auc, probs = 0.75, na.rm = TRUE),
            auc.mean = mean(pt.auc, na.rm = TRUE),
            auc.sd = sd(pt.auc, na.rm = TRUE)) %>%
  mutate(n.pct = paste0('Measurements / Pt:\n1: ',
                        round((n.1 / nrow(brain.oneobs))*100), '% (', n.1,
                        ')\n2: ', round((n.2 / nrow(brain.oneobs))*100), '% (', n.2,
                        ')\n3: ', round((n.3 / nrow(brain.oneobs))*100), '% (', n.3, ')'),
         mean.desc = get.results.string(mean.n, mean.50, mean.25, mean.75, mean.mean, mean.sd),
         min.desc = get.results.string(min.n, min.50, min.25, min.75, min.mean, min.sd),
         max.desc = get.results.string(max.n, max.50, max.25, max.75, max.mean, max.sd),
         delta.desc = get.results.string(delta.n, delta.50, delta.25, delta.75, delta.mean, delta.sd),
         auc.desc = get.results.string(auc.n, auc.50, auc.25, auc.75, auc.mean, auc.sd))

## Set text sizes
histspag.axis.text <- 7
summaryhist.axis.title <- 8
summaryhist.axis.text <- 6

## -- Function to create histograms of summary statistics -- ##
bio.hist <- function(usedata,   ## data set to use
                     xvar,      ## character string - name of variable on X axis
                     xlabel,    ## character string - label for X axis
                     resultstr, ## character string - descriptive statistics
                     resultx){  ## numeric - x value for placement of correlation results
  ## Replace name of biomarker with "xvar" to allow use in ggplot
  names(usedata) <- gsub(xvar, 'xvar', names(usedata))
  
  ggplot(aes(x = xvar), data = usedata) +
    geom_histogram(fill = 'blue', alpha = 0.3) +
    scale_x_continuous(name = xlabel) +
    scale_y_continuous(name = 'Frequency') +
    ## Add descriptive statistics
    annotate(geom = 'text', label = as.character(resultstr),
             x = resultx, y = Inf, vjust = 1.5, hjust = 1, size = 2.5) +
    theme_minimal() +
    theme(axis.title = element_text(size = summaryhist.axis.title),
          axis.text = element_text(size = summaryhist.axis.text))
}

## -- Function to create all plots for a given biomarker -- ##
bio.plots <- function(use.bio){
  ## Select only data needed for plots: id, study day, biomarker specified
  use.daily <- brain.daily[brain.daily$study.day %in% c(1, 3, 5), c('id', 'study.day', use.bio)]
  ## Replace actual biomarker name with generic "use.bio" to use in ggplot
  names(use.daily) <- gsub(use.bio, 'use.bio', names(use.daily))
  
  ## Get summary values, descriptive statistics for specified biomarker
  use.summary <- bio.summary.pt[bio.summary.pt$bio.marker == use.bio,]
  use.desc <- bio.summary[bio.summary$bio.marker == use.bio,]
  
  ## Create histogram of all biomarker values, regardless of study day
  bio.hist.all <- ggplot(aes(x = use.bio), data = use.daily) +
    geom_histogram(alpha = 0.3, fill = 'blue') +
    scale_x_continuous(name = 'Biomarker Value, All Study Days') +
    scale_y_continuous(name = 'Frequency') +
    ## Add number of measurements per patient
    annotate(geom = 'text', label = use.desc$n.pct,
             x = max(use.daily$use.bio, na.rm = TRUE), y = Inf,
             vjust = 1.3, hjust = 1, size = 2.5) +
    ## Add title - biomarker name
    ggtitle(toupper(gsub('\\.imp$', '', use.bio))) +
    theme_minimal() +
    theme(title = element_text(hjust = 0, size = 18, face = 'bold'),
          axis.title = element_text(hjust = 0.5, vjust = 0.5, face = 'plain', size = 10),
          axis.text = element_text(size = histspag.axis.text))
  
  ## Create spaghetti plot of biomarker value by study day per patient
  bio.spaghetti <- ggplot(aes(x = study.day, y = use.bio, group = id), data = use.daily) +
    geom_line(alpha = 0.15, colour = 'blue') +
    scale_x_continuous(breaks = c(1, 3, 5), name = 'Study Day') +
    scale_y_continuous(name = 'Biomarker Value') +
    theme_minimal() +
    theme(axis.text = element_text(size = histspag.axis.text))
  
  ## Create histograms of summary values (mean, min, max, delta, AUC)
  bio.mean.hist <- bio.hist(usedata = use.summary,
                            'pt.mean', 'Patient Mean',
                            use.desc$mean.desc, max(use.summary$pt.mean, na.rm = TRUE))
  bio.min.hist <- bio.hist(usedata = use.summary,
                           'pt.min', 'Patient Minimum',
                           use.desc$min.desc, max(use.summary$pt.min, na.rm = TRUE))
  bio.max.hist <- bio.hist(usedata = use.summary,
                           'pt.max', 'Patient Maximum',
                           use.desc$max.desc, max(use.summary$pt.max, na.rm = TRUE))
  bio.delta.hist <- bio.hist(usedata = use.summary,
                             'pt.delta', 'Patient Change, Last - First',
                             use.desc$delta.desc, max(use.summary$pt.delta, na.rm = TRUE))
  bio.auc.hist <- bio.hist(usedata = use.summary,
                           'pt.auc', 'Patient AUC',
                           use.desc$auc.desc, max(use.summary$pt.auc, na.rm = TRUE))
  
  return(list('total.hist' = bio.hist.all,
              'spaghetti' = bio.spaghetti,
              'mean.hist' = bio.mean.hist,
              'min.hist' = bio.min.hist,
              'max.hist' = bio.max.hist,
              'delta.hist' = bio.delta.hist,
              'auc.hist' = bio.auc.hist))
}

## Create list with one element per biomarker, each element has all plots for that marker
all.bio.plots <- lapply(paste0(biomarker.list, '.imp'), bio.plots)

## -- Function to plot all six plots per biomarker using multiplot() -- ##
multiplot.bio <- function(bio.plot.list){
  multiplot(bio.plot.list$total.hist, bio.plot.list$spaghetti,
            bio.plot.list$mean.hist, bio.plot.list$min.hist, bio.plot.list$max.hist,
            bio.plot.list$delta.hist, bio.plot.list$auc.hist,
            layout = matrix(c(1, 2, 6, 1, 2, 7, 3, 4, 5), ncol = 3))
}

pdf(file = 'biomarker_descplots.pdf', width = 10.5, height = 8, onefile = TRUE)
for(i in 1:length(all.bio.plots)){
  multiplot.bio(all.bio.plots[[i]])
}
dev.off()
```

## Scatterplots and Spearman Correlations between Day 1 Biomarkers
We calculated Spearman correlation coefficients and produced scatterplots of each biomarker measured
on study day 1 vs. all other biomarkers. In the plots, the X-axis is the variable in the column, and
the Y-axis is the variable in the row.

```{r, echo=FALSE, results='asis', fig.width = 10, fig.height = 10}
## -- Scatterplots and correlation statistics for day 1 values -- ##
## Get all day 1 values into a matrix
day1.data <- subset(brain.daily, study.day == 1)[,c(paste0(biomarker.list, '.imp'))]

## -- Get data frame with all unique combinations of biomarkers -- ##
## Use expand.grid to get all unique combinations, but includes duplicates
marker.combs <- expand.grid(names(day1.data), names(day1.data))

## Create new variable that is *sorted* marker1_marker2
marker.combs$unique.comb <- unlist(lapply(1:nrow(marker.combs), FUN = function(i){
  markervals <- c(as.character(marker.combs[i, 'Var1']), as.character(marker.combs[i, 'Var2']))
  paste(sort(markervals), collapse = '_') }))

## Use strsplit to get all sorted combinations of biomarkers, then take all unique values
markercomb.strsplit <- lapply(unique(marker.combs$unique.comb), FUN = strsplit, split = '_')
marker.comb.unique <-
  data.frame(marker1 = unlist(lapply(markercomb.strsplit, FUN = function(x){ x[[1]][1] })),
             marker2 = unlist(lapply(markercomb.strsplit, FUN = function(x){ x[[1]][2] })))
marker.comb.unique <- subset(marker.comb.unique, marker1 != marker2)

## -- Create matrix for multiplot: lower triangle is scatterplots, upper triangle is rho/Ps -- ##
corrplotmat <- matrix(NA, ncol = length(biomarker.list), nrow = length(biomarker.list))
corrplotmat[lower.tri(corrplotmat)] <- 1:sum(lower.tri(corrplotmat))
corrplotmat <- t(corrplotmat)
corrplotmat[lower.tri(corrplotmat)] <- 1:sum(lower.tri(corrplotmat))
corrplotmat[upper.tri(corrplotmat)] <- corrplotmat[upper.tri(corrplotmat)] + sum(lower.tri(corrplotmat))
diag(corrplotmat) <- (sum(!is.na(corrplotmat)) + 1):length(corrplotmat)

## Function to create scatterplot with smoother, given an X and Y biomarker
scattersmooth <- function(xvar, yvar, dataset){
  if(xvar == yvar){
    return(NULL)
  } else{
    usedata <- dataset[,c(xvar, yvar)]
    names(usedata) <- c('xvar', 'yvar')
    
    ggplot(aes(x = xvar, y = yvar), data = usedata) +
      geom_point(size = 1, alpha = 0.2) +
      geom_smooth(alpha = 0.2) +
      theme_minimal() +
      labs(x = NULL, y = NULL) +
      theme(axis.text = element_blank(),
            axis.ticks = element_blank(),
            panel.grid = element_blank(),
            plot.margin=unit(c(0,0,0,0),"mm"))
  }
}

all.scatter <- lapply(1:nrow(marker.comb.unique), FUN = function(i){
  scattersmooth(xvar = marker.comb.unique[i, 1],
                yvar = marker.comb.unique[i, 2],
                dataset = day1.data)
})

## Function to create "plot" with rho, p-value, given an X and Y biomarker
rhop <- function(xvar, yvar, dataset){
  if(xvar == yvar){
    return(NULL)
  } else{
    usedata <- dataset[,c(xvar, yvar)]
    names(usedata) <- c('xvar', 'yvar')
    
    corrstats <- rcorr(as.matrix(usedata[,1:2]), type = 'spearman')
    corrstr <- paste0('N: ', corrstats$n['xvar', 'yvar'],
                      '\nR: ', round.format(corrstats$r['xvar', 'yvar'], dig = 2),
                      '\nP: ', round.format(corrstats$P['xvar', 'yvar'], dig = 3))
    ggplot(aes(x = xvar, y = yvar), data = usedata) +
      annotate(geom = 'text', label = corrstr, size = 3,
               x = mean(usedata$xvar, na.rm = TRUE), y = mean(usedata$yvar, na.rm = TRUE)) +
      theme_minimal() +
      theme(axis.title = element_blank(),
            axis.text = element_blank(),
            axis.ticks = element_blank(),
            panel.grid = element_blank(),
            plot.margin=unit(c(0,0,0,0),"mm"))
  }
}

all.rho <- lapply(1:nrow(marker.comb.unique), FUN = function(i){
  rhop(xvar = marker.comb.unique[i, 1],
       yvar = marker.comb.unique[i, 2],
       dataset = day1.data)
})

## Create "plots" with text for biomarker labels to put on diagonals
bio.plot.fakedata <- data.frame(xval = 1:10, yval = 1:10)
bio.plots <- lapply(biomarker.list, FUN = function(marker){
  ggplot(aes(x = xval, y = yval), data = bio.plot.fakedata) +
      annotate(geom = 'text', label = toupper(marker),
               x = 5.5, y = 5.5, face = 'bold', colour = 'navy', size = 3.7) +
      theme_minimal() +
      theme(axis.title = element_blank(),
            axis.text = element_blank(),
            axis.ticks = element_blank(),
            panel.grid = element_blank(),
            plot.margin=unit(c(0,0,0,0),"mm"))
})

## Create final PDF
# pdf('scattercorr.pdf', width = 10.5, height = 8)
multiplot(plotlist = c(all.scatter, all.rho, bio.plots),
          layout = corrplotmat)
# dev.off()

```

## Hierarchical Clustering Analysis of Day 1 Biomarkers
We performed hierarchical clustering analysis using the `varclus()` function in the `Hmisc` library.
This analysis looks at how closely related each biomarker is to the others, using both a Spearman's
rho (nonparametric correlation statistic) and Hoeffding's D, which looks for non-monotonic
relationships. In neither case did we see relationships strong enough to eliminate biomarkers from
the analysis.

```{r varclus, fig.width = 9, fig.height = 6}
# pdf(file = 'biomarker_varclus.pdf', width = 7, height = 6, onefile = TRUE)
par(mfrow = c(1, 2))
plot(varclus(as.matrix(day1.data))) + title('Spearman rho^2')
plot(varclus(as.matrix(day1.data), similarity = 'hoeffding')) + title("Hoeffding's D")
# dev.off()

```

## Redundancy Analysis
We conducted a redundancy analysis, which looks at how much of each biomarker can be predicted from
the other day 1 biomarkers. IFN-$\gamma$ had the strongest relationship with the other biomarkers, but
even so, only a little over half of its variability can be explained by the combination of the other
biomarkers.

```{r redundancy, fig.width = 8, fig.height = 6}
## Create data set to use for redundancy analyses
brain.daily.use <- brain.daily[brain.daily$study.day %in% c(1, 3, 5),
                               c('id', 'study.day', paste0(biomarker.list, '.imp'))]

## -- Function to print redun() results in ggplot form -- ##
redun.ggplot <- function(redun.obj){
  redun.rsq <- sort(redun.obj$rsq1, decreasing = FALSE)
  redun.data <- data.frame(var = names(redun.rsq),
                           rsq = as.numeric(redun.rsq),
                           index = 1:length(redun.rsq))
  redun.plot <- ggplot(aes(x = rsq, y = index), data = redun.data) +
                  geom_point() +
                  scale_x_continuous(limits = c(0, 1)) +
                  scale_y_continuous(limits = c(1, length(redun.rsq)),
                                     breaks = 1:length(redun.rsq),
                                     labels = redun.data$var) +
                  xlab(expression(R^2)) +
                  ylab('')
  return(redun.plot)
}

## -- Specify R2 cutoff -- ##
r2.min <- 0.6

## Redundancy analyses
allbio.redun <- redun(as.formula(paste('~ ', paste(paste0(biomarker.list, '.imp'), collapse = ' + '))),
                      data = brain.daily.use,
                      nk = 5,
                      r2 = r2.min)
allbio.redun.plot <- redun.ggplot(allbio.redun)

day1bio.redun <- redun(as.formula(paste('~ ', paste(paste0(biomarker.list, '.imp'), collapse = ' + '))),
                       data = subset(brain.daily.use, study.day == 1),
                       nk = 5,
                       r2 = r2.min)
day1bio.redun.plot <- redun.ggplot(day1bio.redun)

## Create plot with both versions on a single panel, sorted by "all days" R2
rsq.order <- data.frame(marker = names(allbio.redun$rsq1),
                        marker.order = order(allbio.redun$rsq1, decreasing = FALSE))

all.redun.rsq <- data.frame(marker = rep(names(sort(allbio.redun$rsq1)), 2))
all.redun.rsq$rsq.val <- c(allbio.redun$rsq1[all.redun.rsq$marker[1:length(allbio.redun$rsq1)]],
                           day1bio.redun$rsq1[all.redun.rsq$marker[1:length(day1bio.redun$rsq1)]])
all.redun.rsq$which.rsq <- factor(c(rep('All study days', length(names(allbio.redun$rsq1))),
                                    (rep('Day 1 only', length(names(allbio.redun$rsq1))))))
all.redun.rsq$plot.order <- rep(1:length(allbio.redun$rsq1), 2)

# pdf(file = 'redundancy.pdf', width = 6, height = 6)
ggplot(aes(x = rsq.val, y = plot.order, colour = which.rsq), data = all.redun.rsq) +
  geom_point(size = 3, alpha = 0.7) +
  scale_colour_discrete(name = 'Days used') +
  scale_x_continuous(limits = 0:1, name = expression(R^2)) +
  scale_y_continuous(name = '',
                     breaks = unique(all.redun.rsq$plot.order),
                     labels = toupper(gsub('\\.imp$', '', all.redun.rsq$marker[1:length(allbio.redun$rsq1)]))) +
  theme_minimal() +
  theme(legend.position = c(1, 0), legend.justification = c(1, 0), axis.ticks = element_blank())
# dev.off()

```

## Comparing Area Under the Curve

For each biomarker and LTCI test/time point, we calculated AUC for nine ROC curves using the 10th,
20th, ..., 90th percentiles as the cutoff for cognitive impairment. For example, in the figure below,
the line for CRP vs. 3-month RBANS is the AUC for ROC curves with CRP vs. [cognitive impairment =
score at or below the 10th percentile of 3-month RBANS]; CRP vs. [cognitive impairment = score at or
below the 20th percentile of 3-month RBANS]; etc. The idea is that lines which have a significant
amount of overlap may have similar relationships with the outcome and therefore may be redundant in
final analysis.

Vertical dashed lines indicate the traditional cutoff for cognitive impairment using each test (<=78
for the RBANS, and <=35 for Trails B).

One note, however, is that some biomarkers may have opposite relationships with
LTCI - for example, higher values of one biomarker may be associated with worse LTCI scores, while
higher values of another biomarker may be associated with better LTCI scores.


```{r auc, fig.width = 10, fig.height = 8}
## Create preliminary data frame with each LTCI variable, day 1 biomarkers
auc.df <- brain.fu %>%
  filter(fu.period %in% c('3 Month', '12 Month')) %>%
  select(id, fu.period, rbans.global.score, trail.b.tscore) %>%
  gather(key = test, value = score, rbans.global.score:trail.b.tscore) %>%
  separate(fu.period, into = c('futime', 'month'), ' ') %>%
  mutate(test = ifelse(test == 'rbans.global.score', 'rbans', 'trailsb'),
         test = paste(test, futime, sep = '.')) %>%
  select(id, test, score) %>%
  spread(key = test, value = score) %>%
  left_join(filter(brain.daily[c('id', 'study.day', paste0(biomarker.list, '.imp'))],
                   study.day == 1),
            by = 'id') %>%
  select(-study.day)

## Function to, for each decile of a test variable, create variable that is
##  1 if score <= decile, 0 otherwise
ltci.imp <- function(ltci.var, quant){
  testval <- quantile(auc.df[,ltci.var], probs = quant, na.rm = TRUE)
  ifelse(is.na(auc.df[,ltci.var]), NA,
  ifelse(auc.df[,ltci.var] > testval, 0, 1))
}

## Add decile impairment variables to initial dataframe
test.vector <- c('rbans.3', 'rbans.12', 'trailsb.3', 'trailsb.12')
ltci.classvars <- do.call(cbind,
                          lapply(test.vector,
                                 FUN = function(testscore){
                                   do.call(cbind,
                                           lapply(seq(0.1, 0.9, 0.1), FUN = function(q){
                                             ltci.imp(ltci.var = testscore,
                                                      quant = q)
                                           }))
                                   }))
colnames(ltci.classvars) <- unlist(lapply(test.vector, FUN = function(testvec){
  paste0(testvec, '.', 1:9)
}))
auc.df <- cbind(auc.df, ltci.classvars)

## Function to get AUC for each combination of biomarker, test, quantile
get.auc <- function(marker, ltci.var){
  unlist(lapply(1:9, FUN = function(quant){
    as.numeric(roc(auc.df[,paste(ltci.var, quant, sep = '.')], auc.df[,marker])$auc)
  }))
}

auc.values <- bind_rows(lapply(paste0(biomarker.list, '.imp'), FUN = function(mark){
  bind_rows(lapply(test.vector, FUN = function(ltci.test){
    data.frame(biomarker = rep(mark, 9),
               ltci.test = rep(ltci.test, 9),
               pctile = paste0(1:9, '0th'),
               auc.value = get.auc(marker = mark, ltci.var = ltci.test))
  }))
}))

auc.values$ltci.test.f <- with(auc.values, {
  factor(ifelse(ltci.test == 'rbans.3', 1,
         ifelse(ltci.test == 'rbans.12', 2,
         ifelse(ltci.test == 'trailsb.3', 3, 4))),
         levels = 1:4,
         labels = c('RBANS, 3 Months', 'RBANS, 12 Months', 'Trails B, 3 Months', 'Trails B, 12 Months')) })

## Get percentile of impairment cutoff for each test/time
pctile.impair.cutoffs <-
  data.frame(ltci.test.f = factor(1:4, labels = levels(auc.values$ltci.test.f)),
             imp.cutoff = c(ecdf(auc.df$rbans.3)(78), ecdf(auc.df$rbans.12)(78),
                            ecdf(auc.df$trailsb.3)(35), ecdf(auc.df$trailsb.12)(35)))

# pdf(file = 'auc_biomarkers.pdf', width = 6, height = 6)
ggplot(aes(x = as.numeric(pctile), y = auc.value, group = biomarker, colour = biomarker),
       data = auc.values) +
  facet_wrap(~ ltci.test.f, nrow = 2) +
  geom_vline(aes(xintercept = imp.cutoff*10),
             data = pctile.impair.cutoffs, alpha = 0.4, linetype = 'dashed') +
  geom_line(alpha = 0.6) +
  geom_point() +
  scale_colour_discrete(name = 'Biomarker') +
  scale_x_continuous(name = 'Percentile', breaks = 1:9, labels = unique(auc.values$pctile)) +
  scale_y_continuous(name = 'Area Under ROC Curve') +
  theme_bw()
# dev.off()

```

```{r corrmarkermeasures}
#### Correlations and scatterplots for each summary measurement for each biomarker; should inform
####  discussion about which summary measures to look at

## -- Plot setup -- ##
## Create vector of summary measure variable names, data frame of all possible combinations (numeric)
summary.measure.list <- setdiff(names(bio.summary.pt), c('id', 'bio.marker', 'pt.n'))
n.summeasures <- length(summary.measure.list)
summeasures.num <- bind_rows(lapply(1:(n.summeasures - 1), FUN = function(i){
  data.frame(var1 = rep(i, n.summeasures - i),
             var2 = (i + 1):n.summeasures)
}))

## -- Create matrix for multiplot: lower triangle is scatterplots, upper triangle is rho/Ps -- ##
sumcorrplotmat <- matrix(NA, ncol = length(summary.measure.list), nrow = length(summary.measure.list))
sumcorrplotmat[lower.tri(sumcorrplotmat)] <- 1:sum(lower.tri(sumcorrplotmat))
sumcorrplotmat <- t(sumcorrplotmat)
sumcorrplotmat[lower.tri(sumcorrplotmat)] <- 1:sum(lower.tri(sumcorrplotmat))
sumcorrplotmat[upper.tri(sumcorrplotmat)] <- sumcorrplotmat[upper.tri(sumcorrplotmat)] + sum(lower.tri(sumcorrplotmat))
diag(sumcorrplotmat) <- (sum(!is.na(sumcorrplotmat)) + 1):length(sumcorrplotmat)

## Add a row at the top for biomarker name plot
sumcorrplotmat <- rbind(c(rep((max(sumcorrplotmat) + 1), 2),
                          seq((max(sumcorrplotmat) + 2), by = 1, length.out = (ncol(sumcorrplotmat) - 2))),
                        sumcorrplotmat)


sumcorr.marker <- function(marker){
  ## Create "plots" with text for biomarker label for title
  sum.bio.labels <- c(toupper(marker), rep('', length(summary.measure.list) - 2))
  sum.bio.plots <- lapply(sum.bio.labels, FUN = function(markertext){
    ggplot(aes(x = xval, y = yval), data = bio.plot.fakedata) +
      annotate(geom = 'text', label = markertext,
               x = 0, y = 0, hjust = 0, vjust = 0, fontface = 'bold', size = 10) +
      scale_x_continuous(expand = c(0, 0), limits = c(0, 10)) +
      scale_y_continuous(expand = c(0, 0), limits = c(0, 5)) +
      theme_minimal() +
      theme(axis.title = element_blank(),
            axis.text = element_blank(),
            axis.ticks = element_blank(),
            panel.grid = element_blank(),
            plot.margin=unit(c(0,0,0,0),"mm"))
  })

  ## Create scatterplots with smoother
  sum.scatter <- lapply(1:nrow(summeasures.num), FUN = function(i){
    scattersmooth(xvar = summary.measure.list[as.numeric(summeasures.num[i, 1])],
                  yvar = summary.measure.list[as.numeric(summeasures.num[i, 2])],
                  dataset = subset(bio.summary.pt, bio.marker == paste0(marker, '.imp')))
  })

  ## Create "plots" with N, rho, p  
  sum.rho <- lapply(1:nrow(summeasures.num), FUN = function(i){
    rhop(xvar = summary.measure.list[as.numeric(summeasures.num[i, 1])],
         yvar = summary.measure.list[as.numeric(summeasures.num[i, 2])],
         dataset = subset(bio.summary.pt, bio.marker == paste0(marker, '.imp')))
  })
  
  ## Create "plots" with text for summary measure to put on diagonals
  sum.plots <- lapply(summary.measure.list, FUN = function(measure){
    ggplot(aes(x = xval, y = yval), data = bio.plot.fakedata) +
      annotate(geom = 'text', label = capitalize(gsub('^pt\\.', '', measure)),
               x = 5.5, y = 5.5, face = 'bold', colour = 'navy', size = 3.7) +
      theme_minimal() +
      theme(axis.title = element_blank(),
            axis.text = element_blank(),
            axis.ticks = element_blank(),
            panel.grid = element_blank(),
            plot.margin=unit(c(0,0,0,0),"mm"))
  })
  
  return(list('scatterplots' = sum.scatter,
              'rhoplots' = sum.rho,
              'measureplots' = sum.plots,
              'titleplots' = sum.bio.plots))
}

bio.summary.plots <- lapply(biomarker.list, FUN = sumcorr.marker)

## Create final PDF
pdf('summaryplots.pdf', width = 9, height = 8)
for(i in 1:length(bio.summary.plots)){
  multiplot(plotlist = c(bio.summary.plots[[i]]$scatterplots,
                         bio.summary.plots[[i]]$rhoplots,
                         bio.summary.plots[[i]]$measureplots,
                         bio.summary.plots[[i]]$titleplots),
            layout = sumcorrplotmat)
}
dev.off()

```
